Rusty Linux: Advances in Rust for Linux Kernel Development

### 2024-06-03 paper.tex
Update on Overleaf.

The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
of breaking into computing systems in cyberspace have come to fruition. The White House Office of
the National Cyber Director (ONCD) released a report calling for the technical community to
proactively reduce the attack surface in cyberspace with a two-pronged approach. First, we need to
address the root cause of many of the most heinous cyber attacks, memory unsafe programming
languages~\cite{United_States_Gov2024-pp}. Second, we need to establish better cybersecurity quality
metrics so we can have a better understanding of the cybersecurity landscape.

@@

Most operating system kernels are predominantly written in the C programming language with bits of
assembly. C has been favored for its low-level capabilities and performance efficiency, crucial for
kernel development. However, the inherent lack of memory safety in C has led to numerous security
vulnerabilities, including buffer overflows and use-after-free errors, which have plagued operation
system development for decades~\cite{noauthor_2019-bf}. Addressing these vulnerabilities is
paramount to enhancing the security and reliability of all operating systems.

Rust, a systems programming language developed by Mozilla, has garnered significant attention for
its strong emphasis on memory safety without sacrificing performance. Rust's ownership model and its
compile-time checks effectively prevent common programming errors that lead to security
vulnerabilities. What makes Rust unique is the fact that it accomplishes these tasks without the use
of a garbage collector, making it an ideal candidate for kernel development where safety and
performance are critical~\cite{}.

The potential integration of Rust into kernel development represents a significant shift in the
landscape of operating system development. This shift prompts a comprehensive evaluation of both the
opportunities and challenges associated with using Rust as such a foundational component of modern
computing. This systematic literature review aims to synthesize existing research and practical
experiences related to the use of Rust in the kernel, providing a detailed understanding of
the current state of this emerging field.

This paper presents a systematic literature review (SLR) focusing on strategies and methodologies
for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe
languages, the operating system kernel. Our study aims to provide a comprehensive overview of
existing research, identify gaps, and suggest future directions in this domain. Through a rigorous
search process, we synthesized relevant studies and extracted key findings to offer insights into
effective approaches for ensuring memory safety when working closely with hardware.

@@

For our research methodology we followed the advice of Kitchenham and Charters~\cite{Stuart2007-cc}
and divided our review into the 3 discrete phases, planning the review, conducting the review, and
reporting the review results. The following sections detail our review process.

@@

Before we start researching we must first confirm the need for a SLR. The recent report released by
ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
detailing the need for research in the domain of memory safety. While the ONCD report detailed a two
pronged approach, for this SLR we will be focusing on a memory safe programming language,
specifically Rust. While there are many modern memory safe programming languages available for
software developers to use Rust is one of the few languages that is feasible to use when developing
operating system kernels due to is lack of runtime and garbage collector. Even the Linux kernel is
in the early stages of adding real support for
Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary
candidate to replace the aging C programming language.

\subsubsection{Research Questions}
\label{sec:researchQuestions}

@@

\begin{itemize}
    \item \textbf{RQ1:} What are the existing approaches and methodologies for implementing
      operating system kernels in Rust?
    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system
      kernel development in terms of throughput, latency, and resource utilization?
    \item \textbf{RQ3:} What are the major challenges, limitations, and lessons learned when
      developing operating system kernels in Rust, particularly in comparison to other languages?
\end{itemize}

\subsubsection{Search Strategy:}

We employed a robust multi-step search strategy across three digital databases, ACM Digital Library,
IEEE Xplore, and Scopus in order to find all the current research that has been done. These
databases cover a wide range of current software engineering and computer science literature. We
leveraged the advanced search features of all three databases to search the titles and abstracts for
keywords using boolean search operators.

\subsubsection{Search criteria:}

We search all three databases with the keywords listed in table~\ref{tab:keywords} between January
1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results Scopus had 44
duplicates that needed to be removed. We then used Google Scholar for forward and backward
snowballing. We choose to look at only the previous 5 years of research in order to capture the
bleeding edge research that is currently being done.

@@

\subsubsection{Inclusion and exclusion criteria:}

Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
inclusion and exclusion criteria based on our research questions outlined in
section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
describe using the Rust programming language for either developing a new kernel, extending an
existing kernel, or authoring drivers. We will include any type of kernel architecture including
monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
the paper is using the Rust programming language in some way. Papers that describe solutions that
reside 100\% in user space will be excluded. Early efforts to use Rust in kernel development
were impacted by changes and updates in both the Rust language and compiler and relied on unstable
features and thus are excluded from this review.

\subsection{Conducting the review}

Once we had our initial collection of papers from the databases we merged all duplicate papers into
one record and then started our first pass which consisted of reading the title and abstract and
either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
completed we were left with 26 potential papers that we needed an in depth reading with an
additional 2 papers found during forward and backward snowballing.

>>>>

The 1995 movie Hackers~\cite{Wikipedia_contributors2024-zr} prescient predictions regarding the ease
of breaking into computing systems in cyberspace have come to fruition. The White House Office of
the National Cyber Director (ONCD) released a report calling for the technical community to
proactively reduce the attack surface in cyberspace with a two-pronged approach~\cite{United_States_Gov2024-pp}. First, we need to
address the root cause of many of the most heinous cyber attacks: memory-unsafe programming
languages. Second, we need to establish better cybersecurity quality
metrics so we can have a better understanding of the cybersecurity landscape.

@@

Most operating system kernels are predominantly written in the C programming language with bits of
assembly. C has been favored for its low-level capabilities and performance efficiency, crucial for
kernel development. However, the inherent lack of memory safety in C has led to numerous security
vulnerabilities, including buffer overflows and use-after-free errors, which have plagued operation
system development for decades~\cite{MSRC_Team2019-bf}. Addressing these vulnerabilities is
paramount to enhancing the security and reliability of all operating systems.

Rust, a systems programming language developed by Mozilla, has garnered significant attention for
its strong emphasis on memory safety without sacrificing performance. Rust's ownership model and its
compile-time checks effectively prevent common programming errors that lead to security
vulnerabilities. What makes Rust unique is the fact that it accomplishes these tasks without the use
of a garbage collector, making it an ideal candidate for kernel development where safety and
performance are critical~\cite{Klabnik2024-id}.

The potential integration of Rust into kernel development represents a significant shift in the
landscape of operating system development. This shift prompts a comprehensive evaluation of both the
opportunities and challenges associated with using Rust as such a foundational component of modern
computing. Therefore, we aim to synthesize existing research and practical
experiences related to the use of Rust in the kernel, providing a detailed understanding of
the current state of this emerging field.

This reflection paper presents a systematic literature review (SLR) focusing on strategies and methodologies
for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe
languages, the operating system kernel. We aim to provide a comprehensive overview of
existing research, identify gaps, and suggest future directions in this domain. Through a rigorous
search process, we synthesized relevant studies and extracted key findings to offer insights into
effective approaches for ensuring memory safety when working closely with hardware.

@@

For our research methodology, we followed the Kitchenham and Charters~\cite{Stuart2007-cc} methodology to conduct SLR in software engineering.
We divided our review into three discrete phases, planning the review, conducting the review, and
reporting the review results. The following sections detail our review process.

@@

To begin our study, we must first confirm the need for a SLR. The recent report released by
ONCD~\cite{United_States_Gov2024-pp} has conveniently done this job for us by compiling a report
detailing the need for research in the domain of memory safety. While the ONCD report detailed a two-pronged approach, for this reflection paper, we will be focusing on a memory-safe programming language,
specifically Rust. While there are many modern memory-safe programming languages available for
software developers to use, Rust is one of the few languages that is feasible to use when developing
operating system kernels due to its lack of runtime and garbage collector. Even the Linux kernel is
in the early stages of adding real support for
Rust~\cite{The_kernel_development_community_undated-iw} thus, we will focus on Rust as a primary
candidate to replace the aging C programming language.

\subsubsection{\textbf{Research Questions}}
\label{sec:researchQuestions}

@@

\begin{itemize}
    \item \textbf{RQ1:} What are the existing approaches for implementing
      operating system kernels in Rust?
    \item \textbf{RQ2:} What are the performance implications of using Rust for operating system
      kernel development?
      %in terms of throughput, latency, and resource utilization?
    \item \textbf{RQ3:} What are the major challenges, limitations, and lessons learned when
      developing operating system kernels in Rust?% particularly in comparison to other languages?
\end{itemize}

\subsection{Data Collection}

\subsubsection{\textbf{Search Strategy.}}

We employed a robust multi-step search strategy across three databases: ACM Digital Library,
IEEE Xplore, and Scopus to find all the current research. 
%These databases cover a wide range of current software engineering and computer science literature. 
We leveraged the advanced search features of all three databases to search the titles and abstracts for
keywords using boolean search operators.

\subsubsection{\textbf{Search criteria.}}

We searched all three databases with the keywords listed in table~\ref{tab:keywords} between January
1, 2019 and April 1, 2024. While both ACM and IEEE had mutually exclusive results, Scopus had 44
duplicates that needed to be removed. We then used Google Scholar for forward and backward
snowballing. We choose to look at only the previous 5 years of research in order to capture the
bleeding-edge research that is currently being done.

@@

\subsubsection{\textbf{Inclusion and exclusion criteria.}}

Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
inclusion and exclusion criteria based on our research questions outlined in
section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
describe using the Rust programming language for either developing a new kernel, extending an
existing kernel, or authoring drivers. We include any type of kernel architecture, including
monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space, as long as
the paper is using the Rust programming language in some way. Papers that describe solutions that
reside 100\% in user space will be excluded. Early efforts to use Rust in kernel development
were impacted by changes and updates in both the Rust language and compiler and relied on unstable
features, which are thus excluded from this review.

\subsection{Data Analysis}

Once we had our initial collection of papers from the databases, we merged all duplicate papers into
one record and then started our first pass, which consisted of reading the title and abstract and
either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
completed we were left with 26 potential papers that we needed an in-depth reading with an
additional 2 papers were found during forward and backward snowballing.
